---
title: "Using rcdk"
author: "Rajarshi Guha"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using rcdk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Given that much of cheminformatics involves mathematical and statistical
modeling of chemical information, R is a natural platform for such work.
There are many cheminformatics applications that will generate useful information
such as descriptors, fingerprints and so on. While one can always
run these applications to generate data that is then imported into R, it can
be convenient to be able to manipulate chemical structures and generate
chemical information with the R environment.

The CDK is a Java library for cheminformatics that supports a wide variety
of cheminformatics functionality ranging from reading molecular file
formats, performing ring perception and armaticity detection to fingerprint
generation and molecular descriptors.

The goal of the `rcdk` package is to allow an R user to access the cheminformatics
functionality of the CDK from within R. While one can use the
rJava package to make direct calls to specific methods in the CDK, from R,
such usage does not usually follow common R idioms. Thus `rcdk` aims to 
 allow users to use the CDK classes and methods in an R-like fashion.

The library is loaded as follows
```{r echo=TRUE}
library(rcdk)
```
The package also provides an example data set, called `bpdata` which contains
277 molecules, in SMILES format and their associated boiling points (BP)
in Kelvin. The `data.frame` has two columns, viz., the SMILES and the BP.
Molecules names are used as row names:
```{r echo=FALSE}
str(bpdata)
```
# Input and Output

Chemical structures come in a variety of formats and the CDK supports
many of them. Many such formats are disk based and these files can be
parsed and loaded by specifying their full paths

```{r, eval=FALSE}
mols <- load.molecules( c('data1.sdf', '/some/path/data2.sdf') )
```

Note that the above function will load any file format that is supported by
the CDK, so there’s no need to specify formats. In addition one can specify a
URL (which should start with `http://`) to specify remote files as well. The
result of this function is a list of molecule objects. The molecule objects
are of class `jobjRef` (provided by the rJava package). As a result,they
are pretty opaque to the user and are really meant to be processed using
methods from the rcdk or rJava packages.

However, since it loads all the molecules from the specified file into a list,
large files can lead to out of memory errors. In such a situtation it is preferable
to iterate over the file, one structure at a time. Currently this behavior
is supported for SDF and SMILES files. An example of such a usage for a large
SD file would be
```{r eval=FALSE}
iter <- iload.molecules('verybig.sdf', type='sdf')
while(hasNext(iter)) {
 mol <- nextElem(iter)
 print(get.property(mol, "cdk:Title"))
}
```

Another common way to obtain molecule objects is by parsing SMILES
strings. The simplest way to do this is
```{r eval=FALSE}
smile <- 'c1ccccc1CC(=O)C(N)CC1CCCCOC1'
mol <- parse.smiles(smile)[[1]]
```
Usage is more efficient when multiple SMILE are supplied, since then a single
SMILES parser object is used to parse all the supplied SMILES.

If you plan on parsing a large number of SMILES, you may run into memory
issues, due to the large size of `IAtomContainer` objects. In such a case, it
can be useful to call the Java and R garbage collectors explicitly at the
appropriate time. In addition it can be useful to explicitly allocate a large
amount of memory for the JVM. For example,
```{r eval=FALSE}
options("java.parameters"=c("-Xmx4000m"))
library(rcdk)
for (smile in smiles) {
    m <- parse.smiles(smile)
    ## perform operations on this molecule
    
    jcall("java/lang/System","V","gc")
    gc()
}
```
Given a list of molecule objects, it is possible to serialize them to a file in
some specified format. Currently, the only output formats are SMILES or
SDF. To write molecules to a disk file in SDF format.
```{r eval=FALSE}
write.molecules(mols, filename='mymols.sdf')
```
By default, if mols is a list of multiple molecules, all of them will be written
to a single SDF file. If this is not desired, you can write each on to individual
files (which are prefixed by the value of filename):
```{r eval=FALSE}
 write.molecules(mols, filename='mymols.sdf', together=FALSE)
 ```
 Finally, we can generate a SMILES representation of a molecule using
 ```{r}
smiles <- c('CCC', 'c1ccccc1', 'CCCC(C)(C)CC(=O)NC')
mols <- parse.smiles(smiles)
get.smiles(mols[[1]])
unlist(lapply(mols, get.smiles))
 ```

# Visualization

The `rcdk` package supports 2D rendering of chemical structures. This can be
used to view the structure of individual molecules or multiple molecules in
a tabular format. It is also possible to view a molecular-data table, where
one of the columns is the 2D image and the remainder can contain data
associated with the molecules.

Due to Java event handling issues on OS X, depictions are handled using an external helper, which means that 
depiction generation can be slower on OS X compared to other platforms. 

Molecule visualization is performed using the `view.molecule.2d` function. For viewing a single molecule or a list of multiple molecules, it is simply
```{r eval=FALSE}
smiles <- c('CCC', 'CCN', 'CCN(C)(C)',
            'c1ccccc1Cc1ccccc1',
            'C1CCC1CC(CN(C)(C))CC(=O)CC')
mols <- parse.smiles(smiles)
view.molecule.2d(mols[[1]])
view.molecule.2d(mols)
```

# Manipulating Molecules

## Adding Information to Molecules

## Atoms and Bonds

## Substructure matching

# Molecular Descriptors

# Fingerprints

# Handling Molecular Formulae